Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-09-03 paper/zhang-cook-laa-langrene-menendez.Rmd
try new web article format

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics 
    url: Melbourne, Australia
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    url: Vienna, Austria
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    url: Melbourne, Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  patricia.menendez@monash.edu 
abstract: A guided tour helps to visualise high-dimensional data by showing low-dimensional projections along a projection pursuit optimisation path. Projection pursuit is a generalisation of principal component analysis, in the sense that different indexes are used to define the interestingness of the projected data. While much work has been done in developing new indexes in the literature, less has been done on understanding the optimisation. Index functions can be noisy, might have multiple local maxima as well as an optimal maximum, and are constrained to generate orthonormal projection frames, which complicates the optimization. In addition, projection pursuit is primarily used for exploratory data analysis, and finding the local maxima is also useful. The guided tour is especially useful for exploration, because it conducts geodesic interpolation connecting steps in the optimisation and shows how the projected data changes as a maxima is approached. This work provides new visual diagnostics for examining a choice of optimisation procedure, based on the provision of a new data object which collects information throughout the optimisation. It has helped to diagnose and fix several problems with projection pursuit guided tour. This work might be useful more broadly for diagnosing optimisers, and comparing their performance. The diagnostics are implemented in the R package \pkg{ferrn}.
keywords: graphics, multivariate, optimization
preamble: >
  \usepackage{mathtools}
  \usepackage[ruled, linesnumbered]{algorithm2e}
output: rticles::rjournal_article
---

@@

```{r toy-pca-animated, fig.cap = "Six frames selected from the animated version of the previous plot. With animation, the progression of the search paths from start to finish is better identified. CRS (green) finishes the optimisation quicker than PD (brown) since there is no further movement for CRS in the sixth frame. The full video of the animation can be found at \\url{https://vimeo.com/504242845}."}
```

An animation is another type of display to show how the search progresses from start to finish in the space. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that CRS finds its end basis quicker than PD since CRS finishes its search in the 5th frame while PD is still making more progress.

@@

```{r toy-tour, fig.cap="Six frames selected from rotating the high-dimensional basis space, along with the same two search paths from Figure \\ref{fig:toy-pca} and \\ref{fig:toy-pca-animated}. The basis space in this example is a 5D unit sphere, on which points (grey) are randomly generated via the CRAN package \\CRANpkg{geozoo}. The full video of the animation can be found at \\url{https://vimeo.com/512885379}."}
```

As mentioned previously, the original $p \times d$ dimension space can be simulated via randomly generated bases in the \pkg{geozoo} [@geozoo] package. While the PCA plot projects the bases from the direction that maximises the variance, the tour plot displays the original high-dimensional space from various directions using animation. Figure \ref{fig:toy-tour} shows some frames from the tour plot of the same two optimisations in its original space.

@@

Both optimisers have found the third variable in the first direction, and the second variable in the second direction. Note however the different orientation of the basis, following from the different sign in the second column. One would expect to see this in the torus plot as the final bases match each other when projected onto one torus circle (due to the same sign in the first column) and are symmetric when projected onto the other (due the sign difference in the second column). In Figure \ref{fig:toy-torus}, this can be seen most clearly in frame 5 where the two circles are rotated into a line from our view.

```{r toy-torus, fig.cap="Six frames selected from rotating the 2D basis space along with two search paths optimised by PD (brown) and CRS (green). The projection problem is a 2D projection with three variables using the holes index. The grey points are randomly generated 2D projection bases in the space and it can be observed that these points form a torus. The full video of the animation can be found at \\url{https://vimeo.com/512882784}."}
```

@@

\bibliography{zhang-cook-laa-langrene-menendez}

>>>>

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics 
    url: Melbourne, Australia
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    url: Vienna, Austria
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    url: Melbourne, Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  patricia.menendez@monash.edu 
abstract: >
  A guided tour helps to visualise high-dimensional data by showing low-dimensional projections along a projection pursuit optimisation path. Projection pursuit is a generalisation of principal component analysis, in the sense that different indexes are used to define the interestingness of the projected data. While much work has been done in developing new indexes in the literature, less has been done on understanding the optimisation. Index functions can be noisy, might have multiple local maxima as well as an optimal maximum, and are constrained to generate orthonormal projection frames, which complicates the optimization. In addition, projection pursuit is primarily used for exploratory data analysis, and finding the local maxima is also useful. The guided tour is especially useful for exploration, because it conducts geodesic interpolation connecting steps in the optimisation and shows how the projected data changes as a maxima is approached. This work provides new visual diagnostics for examining a choice of optimisation procedure, based on the provision of a new data object which collects information throughout the optimisation. It has helped to diagnose and fix several problems with projection pursuit guided tour. This work might be useful more broadly for diagnosing optimisers, and comparing their performance. The diagnostics are implemented in the R package \pkg{ferrn}.
keywords: graphics, multivariate, optimization
preamble: >
  \usepackage{mathtools}
  \usepackage[ruled, linesnumbered]{algorithm2e}
output: 
  rjtools::rjournal_web_article
bibliography: zhang-cook-laa-langrene-menendez.bib
---

@@

An animation is another type of display to show how the search progresses from start to finish in the space. Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\ref{fig:toy-pca-animated-interactive}', '\\ref{fig:toy-pca-animated}'))` shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that CRS finds its end basis quicker than PD since CRS finishes its search in the 5th frame while PD is still making more progress.

```{r toy-pca-animated, fig.cap = "Six frames selected from the animated version of the previous plot. With animation, the progression of the search paths from start to finish is better identified. CRS (green) finishes the optimisation quicker than PD (brown) since there is no further movement for CRS in the sixth frame. The full video of the animation can be found at \\url{https://vimeo.com/504242845}.", eval=knitr::is_latex_output()}
```

```{r toy-pca-animated-interactive, fig.height = 7, fig.width = 7, fig.cap = "this is the interactive version", eval=knitr::is_html_output()}
```

@@

As mentioned previously, the original $p \times d$ dimension space can be simulated via randomly generated bases in the \pkg{geozoo} [@geozoo] package. While the PCA plot projects the bases from the direction that maximises the variance, the tour plot displays the original high-dimensional space from various directions using animation. Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\ref{fig:toy-torus-interactive}', '\\ref{fig:toy-torus}'))` shows some frames from the tour plot of the same two optimisations in its original space.

```{r toy-tour, fig.cap="Six frames selected from rotating the high-dimensional basis space, along with the same two search paths from Figure \\ref{fig:toy-pca} and \\ref{fig:toy-pca-animated}. The basis space in this example is a 5D unit sphere, on which points (grey) are randomly generated via the CRAN package geozoo. The full video of the animation can be found at \\url{https://vimeo.com/512885379}."}
```

@@

Both optimisers have found the third variable in the first direction, and the second variable in the second direction. Note however the different orientation of the basis, following from the different sign in the second column. One would expect to see this in the torus plot as the final bases match each other when projected onto one torus circle (due to the same sign in the first column) and are symmetric when projected onto the other (due the sign difference in the second column). In Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\ref{fig:toy-torus-interactive}', '\\ref{fig:toy-torus}'))`, this can be seen most clearly in frame 5 where the two circles are rotated into a line from our view.

```{r toy-torus, fig.cap="Six frames selected from rotating the 2D basis space along with two search paths optimised by PD (brown) and CRS (green). The projection problem is a 2D projection with three variables using the holes index. The grey points are randomly generated 2D projection bases in the space and it can be observed that these points form a torus. The full video of the animation can be found at \\url{https://vimeo.com/512882784}.", eval = knitr::is_latex_output()}
```

```{r toy-torus-interactive, out.height = "10%", out.width = "10%", fig.cap="Six frames selected from rotating the 2D basis space along with two search paths optimised by PD (brown) and CRS (green). The projection problem is a 2D projection with three variables using the holes index. The grey points are randomly generated 2D projection bases in the space and it can be observed that these points form a torus. The full video of the animation can be found at \\url{https://vimeo.com/512882784}.", eval = knitr::is_html_output()}
```

@@
